---
title: "nuovo modello"
author: "Lorenzo Ricciardulli"
date: "2024-07-10"
output: html_document
---
```{r}
library(corrplot)
library(ggplot2)
library(GGally)
library(leaps)
library(ISLR)
library(readxl)
library(combinat)
library(mvtnorm)
library(coda)
library(corrplot)
library(dplyr)
library(tidyr)
library(rjags)
library(rstan)
library(R2jags)
library(bbricks)
library(LaplacesDemon)
```

# 1) DATA VISUALIZATION AND CLEANING

## 1.1) Data loading and cleaning (dummy variable)

```{r}
data = read.csv("Student_Performance.csv")
data$Extracurricular.Activities = ifelse(data$Extracurricular.Activities == "Yes", 1, 0)
str(data)
```

## 1.2) Data visualization (hists and boxplot)

```{r}
par(mfrow = c(2,3))
hist(data[,1])
hist(data[,2])
hist(data[,4])
hist(data[,5])
hist(data[,6])
```

```{r}
boxplot(data$Performance.Index ~ data$Extracurricular.Activities)
```


# 2) REGRESSION


## 2.1) Samples of beta posterior and tau POST

### 2.1.1) Function
```{r}
############################################
## Linear regression with conjugate prior ##
############################################

#################################################################
## Posterior inference for the linear model (conjugate priors) ##
#################################################################

posterior_linear_conjugate = function(y, X, beta.0, V, a, b, S){
  
  ###########
  ## Input ##
  ###########
  
  # y     : response variable, (n,1) vector
  # X     : (n,p) design matrix, each column is one predictor, column one is the unit vector
  
  # beta.0, V : hyperparameters of the Multivariate Normal prior on beta
  # a, b      : hyperparameters of the Gamma prior on tau
  
  # S : number of MCMC iterations
  
  ############
  ## Output ##
  ############
  
  # out : a list with two elements:
  # beta_post   : (S,p) matrix, each row is a draw from the posterior of beta = (beta_1, ..., beta_p)
  # sigma2_post : (S,1) matrix, each row is a draw from the posterior of sigma2 = tau^{-1}
  
  
  library(mvtnorm)
  
  n = nrow(X)
  p = ncol(X)
  
  ## Compute posterior hyperparameters
  
  V.n    = V + t(X)%*%X
  beta.n = solve(V + t(X)%*%X)%*%(V%*%beta.0 + t(X)%*%y)
  
  a.n = a + n
  b.n = b + t(y)%*%y + beta.0%*%V%*%beta.0 - t(beta.n)%*%(V + t(X)%*%X)%*%beta.n
  
  ## To store samples drawn from the posterior of (beta, sigma2)
  
  beta_post   = matrix(NA, S, p)
  sigma2_post = matrix(NA, S, 1)
  
  ###################
  ## Gibbs sampler ##
  ###################
  
  for(s in 1:S){
    
    #######################
    ## (1) Sample sigma2 ##
    #######################
    
    tau  = rgamma(1, a.n/2, b.n/2)
    sigma2 = 1/tau
    
    ##########################################
    ## (2) Sample beta conditionally on tau ##
    ##########################################
    
    beta = c(rmvnorm(1, beta.n, solve(V.n)/tau))
    
    #########################
    ## Store sampled draws ##
    #########################
    
    beta_post[s,]   = beta
    sigma2_post[s,] = sigma2
    
  }
  
  return(posterior = list(beta_post   = beta_post,
                          sigma2_post = sigma2_post))
  
  
}
```

### 2.1.2) Data preparation for function

```{r}
y = data$Performance.Index ## response variable
data_X = data[,-6]
X = model.matrix(y ~., data_X)
p = ncol(X)
beta.0 = rep(0,p)
V      = diag(0.1, p)
a      = 0.01
b      = 0.01
S = 5000
```

### 2.1.3) Sampling beta comma tau POST

```{r}
out = posterior_linear_conjugate(y = y, X = X, beta.0, V, a, b, S = S)
str(out)
beta_post = out$beta_post
sigma2_post = out$sigma2_post

beta_post_mcmc = as.mcmc(beta_post)
sigma2_post_mcmc = as.mcmc(sigma2_post)

colMeans(beta_post_mcmc)
colMeans(sigma2_post_mcmc)
```
 

## 2.2) Boxplots of beta post mcmc

```{r}
par(mar = c(3,5,1,1))

boxplot(out$beta_post, outline = F, ylim = c(-550,550), ylab = expression(beta), las = 1)
abline(h = 0, col = "blue", lty = "dashed")
out_lm = summary(lm(y ~ X - 1))
points(out_lm$coefficients[,1], col = "blue", pch = 16)

par(mar = c(3,5,1,1))

boxplot(out$beta_post, outline = F, ylim = c(-50,5), ylab = expression(beta), las = 1)
abline(h = 0, col = "blue", lty = "dashed")
out_lm = summary(lm(y ~ X - 1))
points(out_lm$coefficients[,1], col = "blue", pch = 16)
```

## 2.3) Credible intervals

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Sample data for the posterior draws (replace with your actual posterior MCMC samples)
set.seed(123)

# Calculate means and credible intervals for the posterior samples
post.means <- apply(beta_post_mcmc[,1:6], 2, mean)
CI <- apply(beta_post_mcmc[,1:6], 2, function(x) quantile(x, c(0.025, 0.975)))

# Prepare posterior draws for plotting
posterior_draws <- pivot_longer(as.data.frame(beta_post_mcmc[,1:6]), cols = 1:6) %>% 
  arrange(name) %>% 
  mutate(density = "posterior")
colnames(posterior_draws) <- c("parameter", "value", "density")

# Define prior means and standard deviations (conjugate normal prior)
taur <- rgamma(1000, 0.01, 0.01)

# Inspect taur values
summary(taur)

# Adjust prior standard deviations to avoid excessive dispersion
adjusted_taur <- sqrt(taur)
prior_means <- rep(0, 6)  # Assuming zero mean for simplicity
prior_sds <- rep(10, 6)   # Assuming standard deviation of 10 for simplicity

# Generate prior draws
prior_draws <- data.frame(
  parameter = rep(c("V1", "V2", "V3", "V4", "V5", "V6"), each = 1000),
  value = unlist(lapply(1:6, function(i) rnorm(1000, prior_means[i], prior_sds[i] * (1 / adjusted_taur[i])))),
  density = rep("prior", 6 * 1000)
)

# Combine prior and posterior draws
draws <- rbind(posterior_draws, prior_draws)



# Plot the densities
p1 <- ggplot(draws, aes(x = value, colour = density, fill = density)) + 
  geom_density(alpha = 0.2) + 
  xlim(c(-50,50)) +
  theme(aspect.ratio = 1) + 
  theme_bw() + 
  scale_fill_manual(values = c("prior" = "darkgreen", "posterior" = "red")) + 
  scale_color_manual(values = c("prior" = "darkgreen", "posterior" = "red")) + 
  facet_wrap(~parameter, scales = "free") + 
  labs(title = "Prior and Posterior Densities of Coefficients (Conjugate Case)",
       x = "Value", y = "Density")

print(p1)


```



# 4) MODEL / VARIABLE SELECTION

## 4.1) Function for marginal likelihood

```{r}
#######################################
## Computing the marginal likelihood ##
#######################################

marg_like_regr = function(y, X, beta.0, V, a, b){
  
  ###########
  ## Input ##
  ###########
  
  # y     : response variable, (n,1) vector
  # X     : (n,p) design matrix, each column is one predictor, column one is the unit vector
  
  # beta.0, V : hyperparameters of the Multivariate Normal prior on beta
  # a, b      : hyperparameters of the Gamma prior on tau
  
  
  ############
  ## Output ##
  ############
  
  # m : the log-marginal likelihood
  
  X = as.matrix(X)
  V = as.matrix(V)
  
  n = nrow(X)
  p = ncol(X)
  
  ## Compute posterior hyperparameters
  
  V.n    = V + t(X)%*%X
  beta.n = solve(V + t(X)%*%X)%*%(V%*%beta.0 + t(X)%*%y)
  
  a.n = a + n
  b.n = b + t(y)%*%y + beta.0%*%V%*%beta.0 - t(beta.n)%*%(V + t(X)%*%X)%*%beta.n
  
  ## Compute the log-marginal likelihood
  
  m = -n/2*log(2*pi) + a/2*log(b/2) - a.n/2*log(b.n/2) +
        lgamma(a.n/2) - lgamma(a/2) +
          log(det(V))/2 - log(det(V.n))/2
  
  return(m)
  
}
```

## 4.2) Model selection with uniform prior

```{r}
## Find all the possible distinct models with q predictors, q = 1,2,3,...



set = 2:p # labels of variables (intercept, i.e. variable 1 is always included)

combn(set, 1)

1:length(set)

comb = unlist(lapply(1:length(set), combinat::combn, x = set, simplify = FALSE), recursive = FALSE)

tail(comb)

## Obs: the intercept X[,1] must be included in each the 2^9 models

## We add it and also include the model with the intercept only

all.comb = mapply(append, 1, comb, SIMPLIFY = FALSE)

all.models.index = append(1, all.comb)
head(all.models.index)


##################################
## Compute marginal likelihoods ##
##################################

all.log.marg.like = c()

K = length(all.models.index)

beta.0 = rep(0, p)
V      = diag(0.01, p)
a      = 0.01
b      = 0.01

k = 1

for(k in 1:K){
  
  set.k = all.models.index[[k]]
  
  all.log.marg.like[k] = marg_like_regr(y, X[,set.k], beta.0[set.k], V[set.k,set.k], a, b)
  
}

## Obs: we assume the uniform prior on M_k

all.marg.like = exp(all.log.marg.like)

post.model.probs = all.marg.like/sum(all.marg.like)


# alternative: use scale!


scale(all.log.marg.like, center = TRUE, scale = FALSE)

const = max(all.log.marg.like)
all.marg.like = exp(all.log.marg.like - const)  # Usa una costante per la stabilità numerica
post.model.probs = all.marg.like / sum(all.marg.like)

names(post.model.probs) = all.models.index

head(post.model.probs)

round(post.model.probs, 3)

head(sort(post.model.probs, decreasing = TRUE))

par(mar=c(5,12,1,1))
barplot(sort(post.model.probs, decreasing = TRUE)[1:10], horiz = TRUE, las = 2, xlim = c(0,1), col = rainbow(8))

# Compare with BSS and with R squared and adjusted R squared

#####
#####

set = 1:6

lapply(1:length(set), combinat::combn, x = set, simplify = FALSE)
lapply(1:length(set), combinat::combn, x = set, simplify = TRUE)

comb = unlist(lapply(1:length(set), combinat::combn, x = set, simplify = FALSE), recursive = FALSE)
```

## 4.3) Model selection with binomial prior

```{r}
##############################################
## Create a jags object containing the data ##
##############################################

library(R2jags)

jags_data = with(data, list(y = y, X = X, n = length(y), p = ncol(X)))

#################################
## Write likelihood and priors ##
#################################

linear_regr_jags = function(){
  
  
  
  # Gamma prior to tau
  
  tau ~ dgamma(0.01, 0.01)
  
  # Likelihood:
  
  for(i in 1:n){
    
    y[i] ~ dnorm(mu[i], tau)
    mu[i] = (gamma*beta)%*%X[i,]
  }
  
  # Priors:
  
  for(j in 1:p){
    
    beta[j] ~ dnorm(0, tau*0.01)
    
  }
  
  for(j in 1:p){
    
    gamma[j] ~ dbern(w)
    
  }
  
  w ~ dbeta(1, 1)
  
}


#########################################
## Set initial values for beta and tau ##
#########################################

init_values = function(){
  
  list(beta = rep(0, p), gamma = rep(1, p))
  
}


#########################################################################
## Specify for which parameters we want to perform posterior inference ##
#########################################################################

params = c("beta", "gamma")


####################################
## Finally, run the jags function ##
####################################


jags_posterior = jags(data = jags_data,
                      inits = init_values,
                      parameters.to.save = params,
                      model.file = linear_regr_jags,
                      n.chains = 1,
                      n.iter = 1000,
                      n.burnin = 100,
                      n.thin = 1)


out = jags_posterior$BUGSoutput

str(out)

## Extract samples from the posterior of beta and gamma

beta_post  = out$sims.list$beta
gamma_post = out$sims.list$gamma

head(gamma_post)

dim(gamma_post)

S = nrow(gamma_post)

## Estimate the posterior probability of inclusion of each predictor Xj
## i.e. proportion of times gammaj = 1

# Calcola la proporzione di inclusione per ciascun predittore
prob_inclusion <- colMeans(gamma_post)

# Nomina le proporzioni con i nomi delle variabili
names(prob_inclusion) <- c("Intercept", colnames(X)[-1])



# Plot delle proporzioni di inclusione
par(mfrow = c(1, 1), mar = c(5, 5, 2, 2))
barplot_heights <- barplot(prob_inclusion, col = "brown3", ylab = expression(hat(p)[j]), space = 0.4, names.arg = rep("", length(prob_inclusion)))

# Aggiungere le etichette verticali
text(x = barplot_heights, y = par("usr")[3] - 0.01, srt = 90, adj = 1, labels = names(prob_inclusion), xpd = TRUE, cex = 0.7)
```

# 5) PREDICTIONS

```{r}
library(bayesplot)

post_beta_jags = as.matrix(beta_post_mcmc)
post_tau_jags = as.matrix(sigma2_post_mcmc)
# Assuming you have the following objects already:
# post_beta_jags: matrix of posterior samples for coefficients (iterations x predictors)
# post_tau_jags: vector of posterior samples for precision (iterations)
# y: observed values of the dependent variable
# X: design matrix (predictors)

# Number of posterior samples
n_samples <- nrow(beta_post_mcmc)

# Number of observations
n_obs <- length(y)

# Generate posterior predictive samples
posterior_predictive <- matrix(0, n_samples, n_obs)

for (i in 1:n_samples) {
  # Calculate the mean of the predictive distribution
  y_pred_mean <-   X %*% post_beta_jags[i, ] 
  # Calculate the standard deviation (since precision is the reciprocal of variance)
  y_pred_sd <- sqrt(as.vector(post_tau_jags[i]))
  # Generate predictive samples
  posterior_predictive[i, ] <- rnorm(n_obs, y_pred_mean, y_pred_sd)
}

# Convert observed values to a matrix for compatibility with bayesplot
y_matrix <- matrix(y, ncol = length(y))

# Plot Posterior Predictive Checks
color_scheme_set("blue")
ppc_dens_overlay(y, posterior_predictive)
```
## Predictions for us

```{r}
x_new = c(1,8,90,1,7,0)

# Compute mu

mu = beta_post_mcmc%*%x_new # S values of mu

head(mu)

y_pred = rnorm(S, mu, sqrt(sigma2_post_mcmc))

hist(y_pred)
```

